Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises

### 2021-09-23 penguins.Rmd
add distill output format

---
title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises
author:
  - name: Allison M. Horst
    affiliation: Bren School of Environmental Science and Management
    address:
    - University of California, Santa Barbara
    - Santa Barbara, CA 93106-5131
    email:  ahorst@ucsb.edu
  - name: Alison Presmanes Hill
    affiliation: RStudio, PBC
    address:
    - 250 Northern Ave
    - Boston, MA 02210
    email:  alison@rstudio.com
  - name: Kristen B. Gorman
    affiliation: College of Fisheries and Ocean Sciences
    address:
    - University of Alaska Fairbanks 
    - 2150 Koyukuk Drive
    - 245 O’Neill Building
    - Fairbanks, AK 99775-7220
    email:  kbgorman@alaska.edu
abstract: > 
 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 105,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means). 
preamble: |
  % Any extra LaTeX you need in the preamble
output: rticles::rjournal_article
bibiography: penguins.bib
---

@@

>>>>

---
title: Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises
author:
  - name: Allison M. Horst
    affiliation: Bren School of Environmental Science and Management
    address:
    - University of California, Santa Barbara
    - Santa Barbara, CA 93106-5131
    email:  ahorst@ucsb.edu
  - name: Alison Presmanes Hill
    affiliation: RStudio, PBC
    address:
    - 250 Northern Ave
    - Boston, MA 02210
    email:  alison@rstudio.com
  - name: Kristen B. Gorman
    affiliation: College of Fisheries and Ocean Sciences
    address:
    - University of Alaska Fairbanks 
    - 2150 Koyukuk Drive
    - 245 O’Neill Building
    - Fairbanks, AK 99775-7220
    email:  kbgorman@alaska.edu
abstract: > 
 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 105,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means). 
preamble: |
  % Any extra LaTeX you need in the preamble
output: 
  rjdistill::rjournal_web_article: default
  rticles::rjournal_article: default
bibiography: penguins.bib
---

@@

```{r}
library(crosstalk)
shared_penguins <- SharedData$new(penguins, ~species)

simpson_nospecies <- ggplot(data = shared_penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "gray40", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)")

simpson_wspecies <-
  ggplot(data = shared_penguins, aes(x = bill_length_mm, y = bill_depth_mm, group = species)) +
  geom_point(aes(color = species, shape = species), size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(color = species), show.legend = FALSE) +
  scale_color_paletteer_d("colorblindr::OkabeIto") +
  theme_minimal() +
  theme(panel.border = element_rect(fill = NA, color = "gray70")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  guides(color = guide_legend("species"),
         shape = "none")

bscols(widths = c(2,NA),
  list(
    filter_checkbox("species", "Species", shared_penguins, ~species, inline = TRUE)
  ),
  plotly::ggplotly(simpson_wspecies)
)
```
